Numerical Stability of Detonations in White Dwarf Simulations

### 2019-01-23 papers/ignition/paper.tex
Update plots

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
Some simulations of Type Ia supernovae feature thermonuclear detonations that
arise self-consistently during the evolution. It is not clear whether these
detonations would occur in a simulation that fully resolved the burning. In this
study we examine a test detonation problem inspired by collisions of white dwarfs.
This test problem demonstrates that achieving a converged thermonuclear ignition
requires spatial resolution finer than 1 km in the burning region, and that
the simulation as a whole likely needs a base resolution on the order of 1 km.
Current computational resource constraints place this stringent resolution requirement
out of reach for almost all multi-dimensional supernova simulations.
Consequently, contemporary simulations that self-consistently demonstrate
detonations are very likely not resolved and should not be trusted.
\end{abstract}
\keywords{supernovae: general - white dwarfs}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.525]{{{plots/amr_ignition_self-heat}}}
  \caption{Distance from the contact point of the ignition as a function of finest
           spatial resolution. The solid black curve with circle markers represents
           the simulations with uniform resolution (no AMR). The other curves represent
           simulations with a fixed base resolution and varying amounts of AMR.
           \label{fig:self-heat-distance}}
\end{figure}

\autoref{fig:self-heat-distance} shows our main results. The lowest resolution we consider,
256 km, is typical of the early simulations of white dwarf collisions, and demonstrates a
prompt ignition near the contact point. As the (uniform) resolution increases, the ignition
tends to occur earlier and nearer to the contact point. This trend is not physically meaningful:
all simulations with resolution worse than about 1 km represent the same prompt central ignition,
and as the resolution increases, there are grid points physically closer to the center that can ignite.
However, when the resolution is better than 1 km, the situation changes dramatically: the prompt
central ignition does not occur, but rather the ignition is delayed and occurs further from the contact
point. When we have finally converged, the ignition occurs around 1000 km from the contact point,
about 1 second after contact (contrast to less than 0.05 seconds for the simulation with 1 km resolution).
This story contains two important lessons. First, the required resolution for a converged simulation,
less than 100 m, is out of reach of current 3D supernova simulations. Second, the behavior for
resolutions worse than 1 km qualitatively appears to be converged, and one could perhaps
be misled into thinking that there was no reason to try higher resolutions, which is reason
for caution in interpreting reacting hydrodynamics simulations.

>>>>

%==========================================================================
% Abstract
%==========================================================================
\begin{abstract}
Some simulations of Type Ia supernovae feature thermonuclear detonations that
arise self-consistently during the evolution. It is not clear whether these
detonations would occur in a simulation that fully resolved the burning. In this
study we examine a test detonation problem inspired by collisions of white dwarfs.
This test problem demonstrates that achieving a converged thermonuclear ignition
requires spatial resolution much finer than 1 km in the burning region, and that
the simulation as a whole likely needs a base resolution on the order of 1 km or better.
Current computational resource constraints place this stringent resolution requirement
out of reach for almost all multi-dimensional supernova simulations.
Consequently, contemporary simulations that self-consistently demonstrate
detonations are very likely not resolved and should be treated with caution.
\end{abstract}
\keywords{supernovae: general - white dwarfs}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.30]{{{plots/amr_ignition_self-heat}}}
  \caption{Distance from the contact point of the ignition as a function of finest
           spatial resolution. The solid black curve with circle markers represents
           the simulations with uniform resolution (no AMR). The other curves represent
           simulations with a fixed base resolution and varying amounts of AMR.
           \label{fig:self-heat-distance}}
\end{figure}

\autoref{fig:self-heat-distance} shows our main results. The lowest resolution we consider,
256 km, is typical of the early simulations of white dwarf collisions, and demonstrates a
prompt ignition near the contact point. As the (uniform) resolution increases, the ignition
tends to occur earlier and nearer to the contact point. This trend is not physically meaningful:
all simulations with resolution worse than about 1 km represent the same prompt central ignition,
and as the resolution increases, there are grid points physically closer to the center that can ignite.
However, when the resolution is better than 1 km, the situation changes dramatically: the prompt
central ignition does not occur, but rather the ignition is delayed and occurs further from the contact
point. When we have finally converged, the ignition occurs around 900 km from the contact point,
about 1 second after contact (contrast to less than 0.05 seconds for the simulation with 1 km resolution).
This story contains two important lessons. First, the required resolution for a converged simulation,
less than 100 m, is out of reach of current 3D supernova simulations. Second, the behavior for
resolutions worse than 1 km qualitatively appears to be converged, and one could perhaps
be misled into thinking that there was no reason to try higher resolutions, which is reason
for caution in interpreting reacting hydrodynamics simulations.
